INDIGO-DataCloud: a converging evolution of software engineering practices to support European Research e-Infrastructures

### 2019-09-11 paper-jgc.tex
Compacting conclusions

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Both reliability and sustainability of the {\sl INDIGO--DataCloud} software products 
have been primarily improved by acting at the source code level. According to the 
size of the project, which comprised more than 30 products and 200 GitHub repositories, 
this strategy required a remarkable preliminary ground work. On the one hand, the 
definition of a set of SQA requirements -- compiled in the described SQA criteria 
-- that shall be fulfilled prior to release by the {\sl INDIGO--DataCloud} core 
products. On the other hand, the need of setting up the testing infrastructure and
the associated CI/CD pipelines to check that the changes in the source code matched 
those requirements in an automated fashion. As a complement to the above strategy, a 
final validation step combined the deployment of the pre--released software versions 
in the project's preview testbed, and subsequently, most of the products were tested
in selected production environments through the aid of EGI's staged--rollout process.

The most noticeable metric that demonstrates a reliability improvement is represented 
by the ratio between the number of software defects uncovered throughout the pre--release
phase of each of the major releases, {\sl INDIGO--1} and {\sl INDIGO--2}, and the ones 
reported by end users within the EGI e--Infrastructure, where core products were 
deployed. The number of bugs detected in the pre--release phase, filed by the SQA
responsibles of the project, surpassed by a factor of 30 the ones reported by those end 
users, which in absolute terms, this value reached no more than 200 bugs over the
observed timeframe. Considering that these low rate of bugs were uncovered throughout 
the testing phase in the project's preview testbed -- where the new releases were 
deployed in order to validate their integration with other component's production 
versions --, one can infer that the previous phase of code analysis -- running unit and
functional test cases for each minor change -- considerably helped in preventing the 
emergence of a greater quantity of software defects. As it was described, this latter
phase of code analysis included as well the compliance with de--facto style standards, 
improving the readability of the code and, consequently, building on the sustainability
path of the software.

Nevertheless, the individual products of {\sl INDIGO--DataCloud} did not undergo 
a progressive quality improvement throughout the lifespan of the project. Indeed,
unit testing coverage dropped for about 18\% of the total products halfway through
the project. This behavior was aligned with the high-demanding periods of software
development, in particular, the previous months before the second major release 
(INDIGO--2). This fact demonstrates that, even considered as best practice, test 
cases are not commonly written right after, or even before, introducing a new unit
or functionality in the code. High coverage values are generally advisable but they
entail time--consuming work that is often neglected by the code owners. The effort 
spent in testing should be oriented to those sections of the code that are more 
likely to cause negative effects to end users. Consequently, a high coverage value 
does not generally provide a linear benefit with respect to the effort spent -- the
SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
it does not indicate whether the key sections of the code are covered. However, unit
testing is a valuable tool to detect defects at coding time, so whenever the test 
cases accompany the changes done in the code, a high unit testing coverage value is 
a good indicator of reliability in the software.
% ?? (related defect increase?, resulting from the INDIGO-2 release?)

All along the project we have observed that source code review is the cornerstone to
improve the completeness of the unit test cases, as a commitment to strenghten the 
structural quality of the software. Therefore, we move forward in avoiding the 
aforementioned risks of having high coverage values obtained from meaningless test 
cases, which execute a great amount of statements and lines of the code, but using
incomplete test cases. Code reviewers analyse the content of each change in the 
source code and the suitability -- in the above terms -- of the associated tests, 
being the last checkpoint before the change is added in the production branch. Hence 
the vital importance, as mentioned earlier, of providing the tests alongside the code
for every given change. As we have seen, the production branches of 
{\sl INDIGO--DataCloud} source code repositories were protected against direct pushes,
requiring a manual assessment step before approval through peer code review.
% ?? figure with code reviews or sth about PRs, external participation

>>>>

% Comment
%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies? Is there some quantitative evidence of uptime, or bug reports per user, or similar? Or some evidence of speed of defect detection & fixing, which you cite as greatly improved? I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
Both reliability and sustainability of the {\sl INDIGO--DataCloud} software products 
have been primarily improved by acting at the source code level. According to the 
size of the project, which comprised more than 30 products and 200 GitHub repositories, 
this strategy required a remarkable preliminary ground work. On the one hand, the 
definition of a set of SQA requirements -- compiled in the described SQA criteria 
-- ought to be fulfilled by the set of core products before being
released. On the other hand, the setup of a testing infrastructure and
the associated CI/CD pipelines that guarantee that the changes done at the source code
level match
those requirements in an automated fashion. As a final step, the validation of the 
resultant pre--released software versions firstly through integration testing in the
project's preview testbed, and subsequently complemented by the EGI's staged--rollout
process, where the brand new versions are adopted in the first place by candidate
resource providers within the e-Infrastructure -- accessible for user evaluation --, 
before being tagged as production--ready software, and consequently, distributed 
through the EGI channels.

The most noticeable metric that demonstrates a reliability improvement is represented 
by the ratio between the number of software defects uncovered throughout the pre--release
phase of each of the major releases, {\sl INDIGO--1} and {\sl INDIGO--2}, and the ones 
reported by the end users within the EGI e--Infrastructure, where the core products were 
deployed. The number of bugs detected throughout the SQA process here described 
surpassed by a factor of 30 the number of bugs reported by those end users, over a total
of almost 200 bugs in the observed timeframe. Considering that this low rate of bugs -- 
taking into account the amount of software products existing in {\sl INDIGO--DataCloud} -- 
correspond to the outcomes of the integration 
testing phase, it can be inferred that the preceding phase of code analysis -- enforcing 
compliance with a style standard, and running
unit and functional test cases for each minor change -- substantially contributed 
to the robustness of the {\sl INDIGO--DataCloud} software, uncovering defects at earlier 
stages, thus leaving the pre--released software with a low amount of errors.
% graph with GitHub defects/issues in development
% graph showing types of defects detected in the integration phase

Nevertheless, the individual products of {\sl INDIGO--DataCloud} did not undergo 
a progressive quality improvement throughout its lifespan. Indeed,
unit testing coverage dropped for about 18\% of the total products halfway through
the project. This behavior was aligned with the high-demanding periods of software
development, in particular, the previous months before the second major release 
(INDIGO--2). This fact demonstrates that, even considered as best practice, test 
cases are not commonly written right after, or even before, introducing a new unit
or functionality in the code. High coverage values are generally advisable but they
entail time--consuming work that is often neglected by the code owners. Rather than
requiring a high coverage value, our experience showed that increasing the coverage
of those sections in the code that are more likely to cause negative effects to end
users is not only a more effective approach to improve software reliability but it 
also keeps developers motivated in the task of test coding, as it brings more 
noticeable benefits in the software operation by end users.
% ?? (related defect increase?, resulting from the INDIGO-2 release?)

%The effort 
%spent in testing should be oriented to those sections of the code that are more 
%likely to cause negative effects to end users. Consequently, a high coverage value 
%does not generally provide a linear benefit with respect to the effort spent -- the
%SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
%it does not indicate whether the key sections of the code are covered. However, unit
%testing is a valuable tool to detect defects at coding time, so whenever the test 
%cases accompany the changes done in the code, a high unit testing coverage value is 
%a good indicator of reliability in the software.

All along the project we have observed that source code review is the cornerstone to
improve the effectiveness of the unit test cases. {\sl INDIGO--DataCloud} source code 
repositories are protected against direct pushes so that the code review stage always
takes place. Code reviewers analyse the content of each change in the source code and
the suitability -- in the above terms -- of the associated tests, being the last
checkpoint before the change is added into the production branch. Hence the vital
importance of providing the tests alongside the code for every given change.
% ?? figure with code reviews or sth about PRs, external participation
